Writing Trash Bin

### 2024-03-18 paper.qmd
rewrote contextual information section

The idea that inference is related to uncertainty pops up in most discussions of uncertainty, however, due to the now *implicit* understanding we are performing inferential statistics, it is often brought up as a *task* or *goal* dependence. This is mentioned both by authors at a specific stage of an analysis, and by authors looking at an entire field or at all stages of the analysis pipeline. @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects depends on the relationship we seek to identify with our data. @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer. @Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. These examples show that authors that discuss uncertainty believe it is there is an important relationship between uncertainty and what we decide to keep in our analysis and throw away, i.e. the task or goal. The task or goal is, in essence, the *statistic you are drawing inference on*. Were this not true, we would have no reason to differentiate between a prediction interval and a sampling distribution because they could both be considered "uncertainty about the mean", however they are different because their inferential goals are different.

@@

### Contextual Information
Some uncertainty work, specifically uncertainty visualisation, focus on the context surrounding the uncertainty, such as the spatial or temporal nature of the estimate. While other discussions of uncertainty may consider this distinction more relevant to their work, here we will consider this contextual information. Contextual information is important for interpretation and understanding of graphics, but it does not in of itself generate uncertainty. Its relevance to the field of "uncertainty visualisation" is through the percieved lack of available channels. For example, a choropleth map has already used colour to represent an estimate, and the concept of space to provide contextual information, there is little remaining room tany spatial

This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty. Operations on data, inherrent randomness, and assumptions that create bias generate uncertainty, which is what we are talking about here. A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data.

Papers that stress the importance of spatial or temporal contexts may be conflating contextual information with uncertainty. @fig-spatial depicts the role of contextual information in graphics, as something that helps us understand the information in the statistical graphic. When we create statistical models and quantify uncertainty, the spatial or temporal aspects are often boilled away [@Otsuka2023], and the only element of it that remains is the inexchangeability of our data. For this reason, there realistically is no such thing as "spatial" uncertainty in the same sense that there is no such thing as "photograph" uncertainty or "coin toss" uncertainty or "baby birth" uncertainty. Maps and photographs boil down to the same statistical objects in the same way that coin tosses and baby births boil down to the same statistical objects. Once we have removed the physical nature these objects are near identical when considered statistically. Of course the nature of them can change based on what we choose to observe (i.e. if we consider a coin landing on its side to be an outcome or if we consider intersex babies as an outcome in births) but this means the contextual information of our data, e.g. whether or not it is spatial or temporal, is only of importance at two stages of the data analysis pipeline. It is first important when we *boil it down* to our data through *observation* and again when we *map our statistical information back to its real world information* to *communicate* it. The only way the context uniquely contributes to uncertainty is in the "boil down" stage, however spatial or temporal uncertainty never refers to that specific cause of uncertainty. @Kinkeldey2014 discussed how uncertainty  can be represented by three components attribute (what) position (where) and temporal (when) and that studied typically deal with uncertainty around attribute but rarely position and time, however it is never specified what considerations should cause attribute uncertainty to be different to position or temporal uncertainty. @Kay2016 did an experiment that showed uncertainty around bus arrival times, however the visualisation used in the experiment, shown in @fig-bustime, is indistinguishable from most work that would be considered "attribute" uncertainty.

@@

It is typically in reference to measurement error, data aggregation, statistical modelling, or other steps that occur *after* the boil down stage, and are therefore not unique to spatial or temporal data. While the spatial or temporal features *may* become relevant at the final communication stage, however this has little do with quantifying uncertainty, and the difficulties in communicating a complicated statistical model through a complicated map is already well discussed in the visualisation literature, and in later sections we will discuss how this concept is related to uncertainty visualisation.

This does not mean we are saying spatial and temporal information is never of importance. Spatial information is *highly relevant* when we translate from the abstract domain back into real world for communication. This position is best explained by @Leland2005 in *The Grammar of Graphics*:

> "Geography is anchored in real space-time and statistics in abstract dimensions. This is a distinction along a continuum rather than a sharp break... but this difference in focus clearly means that a system optimized to handle geography will not be graceful when dealing with statistical graphics."

This paper focuses on uncertainty, specifically uncertainty that is still in the the abstract *statistics* domain. Since we are still in the abstract domain, things that are anchored in real space-time create confusion that is evident in the literature. Spatial and temporal uncertainty papers limit how they communicate their uncertainty because they are considering what can be visualised inside the context provided by maps and, not what they should *try* to visualise based on the uncertainty abstraction. Additionally, since uncertainty is defined in the abstract domain, of the issues faced by spatial uncertainty visualisations are likely shared by other graphics that only differ in contextual information.

@@

@@

The class of "uncertainty visualisation" does not seem to have any rules for what belongs in the group and what makes it different to a "typical" visualisation is rarely discussed. Instead one of two general motivations for the field are provided in the subtext.  
1) Uncertainty is fundamentally differently to other variables due to psychological heuristics involved in the interpretation of uncertainty (e.g. risk aversion).
2) Uncertainty is not of interest in of itself and is typically layered on at the end of a the visualisation pipeline, so uncertainty visualisation is the field of adding error information into already established graphics.  
Neither of these motivations are remotely relevant when we are evaluating the simple task of value extraction. The psychological effects of uncertainty are only relevant for decision making and have nothing to do with value abstraction, and the information overload issue is not specific to uncertainty and is simply an example of a larger dimensionality issue.
Therefore, these explanations for the difference fall short of explaining *why* uncertainty should be treated any differently than any other variable.

There is also the common view that uncertainty as a dimensionality problem. This perspective is not quite right and the literature seems to be about 50/50 on whether uncertainty is a psychological or dimensionality problem. @Kinkeldey2014 also identified this issue when they highlighted that the literature makes it unclear if uncertainty is a variable in of itself, something that should be interpreted with the main variable, or if it is metadata. It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen". The phenomena of inattentional blindness shows that there is no perception without attention and it is powerful enough that participants can fail to be aware of random objects appearing on a screen or a gorilla walking through a basketball game [@simons1999gorillas; @mack2003inattentional]. Therefore if information is mapped to graphical elements that are so low on hierarchy they can be ignored, they might as well not be there at all. The people who don't plot uncertainty because they think its unimportant and the people who relegate uncertainty to the lowest ranking aesthetic on the list are both saying the same thing, "I think the uncertainty is unimportant". Including uncertainty is worth very little if no attention is left to see it. This does not mean we cannot put a *large* amount of information in a graphic. Glyph maps can be used to depict the multi-dimensional information in spatial-temporal data by mapping line plots to map locations, but we still need to decide what information is important. The map can either present trends in the global or local variance depending on whether or not the line plot is scaled globally or locally [@Wickham2012], however smaller details are almost impossible to convey. No matter how much information we try and put in a plot, there will always be only a handful of key takeaways.

>>>>

The idea that inference is related to uncertainty pops up in most discussions of uncertainty, however, due to the now *implicit* understanding we are performing inferential statistics, it is often brought up as a *task* or *goal* dependence. This is mentioned both by authors at a specific stage of an analysis, and by authors looking at an entire field or at all stages of the analysis pipeline. @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data. @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer. @Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. These examples show that authors that discuss uncertainty believe it is there is an important relationship between uncertainty and what we decide to keep in our analysis and throw away, i.e. the task or goal. The task or goal is, in essence, the *statistic you are drawing inference on*. Were this not true, we would have no reason to differentiate between a prediction interval and a sampling distribution because they could both be considered "uncertainty about the mean", however they are different because their inferential goals are different.

@@

### Contextual Information
One of the key components of uncertainty and inference is the noise and signal dichotomy. When @Otsuka2023 discusses the information we boil away to create a "probabilistic kind" he uses the example of a coin flip and the sex of baby births as two things that are *very* different in reality, but it is only through the lense of a Bernoulli distribution that we can view these two things as the same. The fact is that this boiling down, of a real world event into data, is not something that happens independent of the statisticians, but is just as much as product of our interests (the particular relationship we hope to draw inference on) as it is a product of the event's natural properties. This is expressed simply in the text with an extension to the coin flipping example:

> "Recall that we modeled a coin toss by a Bernoulli distribution. That was because we supposed the coin will land either heads or tails and ignored the possibility of its landing on its edge. If we are to take the latter possibility into account, we should use a multinational distribution with three categories. As this example shows, the question of which statistical model we should use is not just determined by the physical nature of the object under study, but rather depends on the modeler’s interests and intentions." @Otsuka2023

With this in mind, let us turn to the concept of "contextual" information. For the purposes of this review, we will consider contexual information to be anything that makes it easy for the audience to connect the data from a probablistic kind to its real world parallel, but is largely irrelevant in the abstract probability space. A key example of this is spatial information, such as borders, road locations, terrains, etc. Spatial information is *highly relevant* when we translate from the abstract domain back into real world for communication. It can help provide reasons for *why* there may be specific characteristics in our data, however it is not required information for the model. @fig-spatial shows the proces of boilling information down into data, visualising it, and including contextual information to improve our scientific reasoning. The concept that contextual information is not the same as statistical information is also highlighted in *The Grammar of Graphics* which states:

> "Geography is anchored in real space-time and statistics in abstract dimensions. This is a distinction along a continuum rather than a sharp break... but this difference in focus clearly means that a system optimized to handle geography will not be graceful when dealing with statistical graphics." @Leland2005

For the remaining of this paper, we are not going to differentiate uncertainty visualisation using contextual information, despite the fact that it is common in the field. Our reasoning for this is that organising the field according to contextual information, promotes the idea that uncertainty is a concept specific to data qualities and ignores the role of inferential statistics.

While other discussions of uncertainty may consider this distinction more relevant to their work, here we will consider this contextual information. Contextual information is important for interpretation and understanding of graphics, but it does not in of itself generate uncertainty. Its relevance to the field of "uncertainty visualisation" is through the percieved lack of available channels. For example, a choropleth map has already used colour to represent an estimate, and position to represent spatial information, which is sometimes relevant to the probabilistic kind, and other times it is simply contextual information (such as when the modelling is done by using a state as a category rather than its geographic information). These two cases, when the spatial information is model relevant and when it is contextual, are rarely separated. It can lead authors to believe the spatial information is always of the highest importance, and thus it is always given the position axis, even when it is not, especially with respect to the uncertainty in a model. A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data.

Similar to the confusing definition of uncertainty, the constant conflation of statistical and contextual information also creates a wide range of problems. @Kinkeldey2014 discussed how uncertainty  can be represented by three components attribute (what) position (where) and temporal (when) and that studied typically deal with uncertainty around attribute but rarely position and time, however it is never specified what considerations should cause attribute uncertainty to be different to position or temporal uncertainty. @Kay2016 did an experiment that showed uncertainty around bus arrival times, however the visualisation used in the experiment, shown in @fig-bustime, is indistinguishable from most work that would be considered "attribute" uncertainty.

@@

This paper focuses on uncertainty, specifically uncertainty that is still in the the abstract *statistics* domain. Spatial and temporal uncertainty papers limit how they communicate their uncertainty because they are considering what can be visualised inside the context provided by maps and, not what they should *try* to visualise based on the uncertainty abstraction. When we view a spatial-temporal visualisation through the lense of trying to visualise the distribution of $P(X|longitude, latitude, time)$ there is no reason this is any different from any other high dimensional visualisation problem.

@@

The focus on contextual information has also created a rift in exactly *why* uncertainty visualisation is a field of interest. The two most common justifications are:
1) Uncertainty is fundamentally differently to other variables due to psychological heuristics involved in the interpretation of uncertainty (e.g. risk aversion).
2) Uncertainty is not of interest in of itself and is typically layered on at the end of a the visualisation pipeline, so uncertainty visualisation is the field of adding error information into already established graphics and *thus* uncertainty information is a high dimensional visualisation problem.  
Papers will often discuss uncertainty in relation to one of these motivating reasons, the evaluation experiments motivated by (1) often perform different visualisations of PDFs, while the papers motivated by (2) will focus on trying to impute uncertainty as error within one of the existing channels, such as colour (*Cite Gap 11*). These dual motivations that co-exist with uncertainty visualisation authors rarely mentioning the others, create confusion in the field as to its purpose. @Kinkeldey2014 also identified this issue when they highlighted that the literature makes it unclear if uncertainty is a variable in of itself, something that should be interpreted with the main variable, or if it is metadata.

For these reasons, the implication that uncertainty is not inference related, the barriers that cause authors to ignore existing visualisation methods, and the confusion created in the motivation of uncertainty visualization as a whole, we do not believe it is fruitful to continue separating the field of uncertainty visualisation according to its contextual information. Therefore, the rest of this review will not separate uncertainty visualisations based on this information.

@@

The class of "uncertainty visualisation" does not seem to have any rules for what belongs in the group and what makes it different to a "typical" visualisation is rarely discussed. *removed bit about two motivations* Neither of these motivations are remotely relevant when we are evaluating the simple task of value extraction. The psychological effects of uncertainty are only relevant for decision making and have nothing to do with value abstraction, and the information overload issue is not specific to uncertainty and is simply an example of a larger dimensionality issue.
Therefore, these explanations for the difference fall short of explaining *why* uncertainty should be treated any differently than any other variable.

There is also the common view that uncertainty as a dimensionality problem. This perspective is not quite right and the literature seems to be about 50/50 on whether uncertainty is a psychological or dimensionality problem.  It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen". The phenomena of inattentional blindness shows that there is no perception without attention and it is powerful enough that participants can fail to be aware of random objects appearing on a screen or a gorilla walking through a basketball game [@simons1999gorillas; @mack2003inattentional]. Therefore if information is mapped to graphical elements that are so low on hierarchy they can be ignored, they might as well not be there at all. The people who don't plot uncertainty because they think its unimportant and the people who relegate uncertainty to the lowest ranking aesthetic on the list are both saying the same thing, "I think the uncertainty is unimportant". Including uncertainty is worth very little if no attention is left to see it. This does not mean we cannot put a *large* amount of information in a graphic. Glyph maps can be used to depict the multi-dimensional information in spatial-temporal data by mapping line plots to map locations, but we still need to decide what information is important. The map can either present trends in the global or local variance depending on whether or not the line plot is scaled globally or locally [@Wickham2012], however smaller details are almost impossible to convey. No matter how much information we try and put in a plot, there will always be only a handful of key takeaways.

### 2024-03-18 writingdumpster.qmd
rewrote contextual information section

>>>>

Cite Gap 11: Papers will often discuss uncertainty in relation to one of these motivating reasons, the evaluation experiments motivated by (1) often perform different visualisations of PDFs, while the papers motivated by (2) will focus on trying to impute uncertainty as error within one of the existing channels, such as colour
